---
title: "Homework 5"
author: "Ethan Marburger"
format: html
editor: visual
---

## Homework 5

Calling appropriate packages 

```{r}
#| echo: false
library(tidyverse)
library(readr)
```

### Task 1: Read in the Data and Modify 

1) Reading in and merging student data files using local paths

```{r}
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)

d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d3)) # 382 students
d3
```

2) read in and combine the data using functions from the tidyverse

```{r}
# Reading in data
student_mat <- read_csv2("student-mat.csv")
student_por <- read_csv2("student-por.csv")
# Inner join on same variables as Q1
error_inner_join <- inner_join(student_mat, student_por,
                               by = c("school", "sex", "age", "address", 
                               "famsize", "Pstatus", "Medu", "Fedu", "Mjob", 
                               "Fjob", "reason", "nursery", "internet"))
print(error_inner_join)
```

 # issue with inner join = more columns were returned then specified
 
 3) Use an inner_join() on all variables other than G1, G2, G3, paid, and absences.
 
```{r}
# Removing undesired columns
student_mat_remove <- student_mat |>
  select(-G1, -G2, -G3, -paid, -absences)
student_por_remove <- student_por |>
  select(-G1, -G2, -G3, -paid, -absences)

# Inner join
student_join <- inner_join(student_mat_remove, student_por_remove)
print(student_join)
```

4) For the math data, Portuguese, and combined data, choose four categorical variables you are interested in and convert those into factor variables in each tibble. 

Mutated factors = sex, guardian, activities, and internet.

```{r}
# Converting student_mat_remove categorical variables to factors
student_mat_remove <- student_mat_remove |>
 mutate(sex = as.factor(sex),
        guardian = as.factor(guardian),
        activities = as.factor(activities),
        internet = as.factor(internet))

# Converting student_por_remove categorical variables to factors
student_por_remove <- student_por_remove |>
 mutate(sex = as.factor(sex),
        guardian = as.factor(guardian),
        activities = as.factor(activities),
        internet = as.factor(internet))

# Converting student_join categorical variables to factors
student_join <- student_join |>
 mutate(sex = as.factor(sex),
        guardian = as.factor(guardian),
        activities = as.factor(activities),
        internet = as.factor(internet))

str(student_join)
```

### Task 2:  Summarize the Data (Very Basic EDA)

#### Categorical Variables

1a) One-way contingency table

```{r}
table(student_join$sex, useNA = "always")
```
What's Interesting?: More female students then male.

Number interpretation: There are 146 males in our student_join tibble

1b) two-way contingency table

```{r}
table(student_join$sex, student_join$internet, useNA = "always")
```

What's Interesting?: The amount of students without internet access.

Interpretation: There are 31 females without access to the internet.

1c) three-way contingency table

```{r}
table(student_join$sex, student_join$guardian, student_join$internet)
```

What's Interesting?: Within these two schools, students whose guardian status is "other" all have internet access.

Interpretation: There are 109 females with a mother and internet access

2a) Create a conditional two-way table using table(). Subset the data with filter()

```{r}
filter_student_join <- student_join |>
  drop_na(sex, internet) |>
  group_by(sex, internet) |>
  filter(sex == "F")

table(filter_student_join$sex, filter_student_join$internet, useNA = "always")
```

What's Interesting?: The amount of female students without internet access.

2b) Create a conditional two-way table using table(). Subset a three-way 
contingency table into a two-way table

```{r}
three_way_tbl <- table(student_join$sex, student_join$guardian, 
                       student_join$internet, useNA = "always")

# Subsetting three way table on access to internet
two_way_tbl_subset <- three_way_tbl[, , "yes"]
print(two_way_tbl_subset)
```

What's Interesting?: There are female students with guardian as "father" with internet access than male students.

3) Create a two-way contingency table using group_by() and summarize() from dplyr. Then use pivot_wider() to make the result look more like the output from table()
```{r}
cat_student_join_plot <- student_join |>
  drop_na(sex, guardian) |>
  group_by(sex, guardian) |>
  summarize(count =n())|>
  pivot_wider(names_from = guardian, values_from = count)

print(cat_student_join_plot)
```

What's Interesting?: Number of female and male students with "father" as guardian is the same.

4) Create a stacked bar graph and a side-by-side bar graph. Give relevant x and y labels, and a title for the plots.

```{r}
# Stacked bar graph
g <- ggplot(data = student_join |> drop_na(sex), 
            aes(x = sex, fill = internet))
g + geom_bar() + 
  labs(x = "Sex", title = "Internet Access by Sex") +
  scale_fill_discrete("Internet") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: More females then males.

```{r}
# Side-by-side bar graph
g + geom_bar(position = "dodge") + 
  labs(x = "Sex", title = "Internet Access by Sex") +
  scale_fill_discrete("Internet") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: More females then males, but side-by-side.

#### Numerical Variables

Reading in data with numerical variables

```{r}
student_math <- read_csv2("student-mat.csv")
student_port <- read_csv2("student-por.csv")
student_data <- inner_join(student_math, student_port, 
                           by = c("sex", "guardian", "internet", "age",
                                  "absences", "G1", "G2", "G3"))

# Keeping numerical variables and a few categorical variables 
student_data <- student_data |>
  select(sex, guardian, internet, age, absences, G1, G2, G3)

print(student_data)
```

What's Interesting?: Some students have zero absences.

1) Find measures of center and spread for three of these variables 
(including at least one G3 variable)

```{r}
# Measure of center and spread
student_data |>
 select(age, absences, G3) |>
 drop_na(age, absences, G3) |>
 summarize(across(c(age, absences, G3), .fns = list("mean" = mean, 
                                      "median" = median, 
                                      "var" = var, 
                                      "sd" = sd, 
                                      "IQR" = IQR), .names = "{.fn}_{.col}"))
```

What's Interesting?: Mean number of absences is 2.

```{r}
# Repeating above, subsetting in a meaningful way
student_data |>
 select(age, absences, G3) |>
 drop_na(age, absences, G3) |>
 filter(age > "15", absences < "3", G3 >= "12") |>
 summarize(across(c(age, absences, G3), .fns = list("mean" = mean, 
                                      "median" = median, 
                                      "var" = var, 
                                      "sd" = sd, 
                                      "IQR" = IQR), .names = "{.fn}_{.col}"))
```

What's Interesting?: When subset by the these criteria, the mean number of absences drops to 0.667.

2) Find measures of center and spread across a single grouping variable for three of these variables (including a G3 variable as one of them)

```{r}
student_data |>
group_by(sex) |> # grouping variable
 select(age, absences, G3) |>
 drop_na(age, absences, G3) |>
 filter(age > "15", absences < "3", G3 >= "12") |>
 summarize(across(c(age, absences, G3), .fns = list("mean" = mean, 
                                      "median" = median, 
                                      "var" = var, 
                                      "sd" = sd, 
                                      "IQR" = IQR), .names = "{.fn}_{.col}"))
```

What's Interesting?: Male students don't meet subset criteria.

3) Find measures of center and spread across a two grouping variable for three of these variables (including a G3 variable as one of them)

```{r}
student_data |>
group_by(sex, internet) |> # grouping variables
 select(age, absences, G3) |>
 drop_na(age, absences, G3) |>
 summarize(across(c(age, absences, G3), .fns = list("mean" = mean, 
                                      "median" = median, 
                                      "var" = var, 
                                      "sd" = sd, 
                                      "IQR" = IQR), .names = "{.fn}_{.col}"))
```

What's Interesting?: The mean absences for males without internet access is zero.

4) Create a correlation matrix between all of the numeric variables

```{r}
cor(student_data[4:8])
```
What's Interesting?: absences is positively correlated with age.

5) Create a histogram, kernel density plot, and boxplot for two of the numeric variables across one of the categorical variables

```{r}
# Kernel Density 1: G1 across sex
KD1 <- ggplot(student_data |> drop_na(G1, sex), aes(x = G1))
KD1 + geom_density(alpha = 0.4, aes(fill = sex)) +
  labs(x = "First Period Grade", title = "Kernel Density: First Period Grade by Sex") +
  scale_fill_discrete("Sex") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: The difference in students in first period scoring 10 by sex.

```{r}
# Kernel Density 2: G2 across sex
KD2 <- ggplot(student_data |> drop_na(G2, sex), aes(x = G2))
KD2 + geom_density(kernel = "gaussian", alpha = 0.4, aes(fill = sex)) +
  labs(x = "Second Period Grade", title = "Kernel Density: Second Period Grade by Sex") +
  scale_fill_discrete("Sex") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: The difference in students in second period scoring ~8.5 by sex.

```{r}
# Histogram 1: G1 across sex
H1 <- ggplot(student_data |> drop_na(G1, sex), aes(x = G1))
H1 + geom_histogram(binwidth = 0.5, alpha = 1, aes(fill = sex)) +
  labs(x = "First Period Grade", title = "Histogram: First Period Grade by Sex") +
  scale_fill_discrete("Sex") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: The only score to be achieved by both sexes was thirteen.

```{r}
# Histogram 2: G2 across sex
H2 <- ggplot(student_data |> drop_na(G2, sex), aes(x = G2))
H2 + geom_histogram(binwidth = 0.75, alpha = 1, aes(fill = sex)) +
  labs(x = "Second Period Grade", title = "Histogram: Second Period Grade by Sex") +
  scale_fill_discrete("Sex") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: Only one male scored a zero in second period.

```{r}
# Box Plot 1: G1 across sex
bp1 <- ggplot(student_data |> drop_na(G1, sex))
bp1 + geom_boxplot(aes(x = G1, y = sex, fill = sex)) +
  labs(x = "First Period Grade", title = "Box Plot: First Period Grade by Sex") +
  scale_fill_discrete("Sex") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: The difference in error bars between the sexes.

```{r}
# Box Plot 2: G2 across sex
bp1 <- ggplot(student_data |> drop_na(G2, sex))
bp1 + geom_boxplot(aes(x = G2, y = sex, fill = sex)) +
  labs(x = "Second Period Grade", title = "Box Plot: Second Period Grade by Sex") +
  scale_fill_discrete("Sex") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: In the second period, male score are spread across the board while female scores are concentrated around ten.

6) Create two scatterplots relating a G3 variable to other numeric variables (put G3 on the y-axis).

```{r}
# Scatter Plot 1: G3 and age across sex
sp1 <- ggplot(student_data |> drop_na(G3, age), aes(x = age, y = G3, color = sex))
sp1 + geom_point() +
  labs(x = "Student Age", y = "Third Period Grades", 
       title = "Scatter Plot: Third Period Grade by Age") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: A single 16 year old male scored a zero in third period.

```{r}
# Scatter Plot 2: G3 and absences across sex
sp1 <- ggplot(student_data |> drop_na(G3, absences), aes(x = absences, y = G3, color = sex))
sp1 + geom_point() +
  labs(x = "Absences", y = "Third Period Grades", 
       title = "Scatter Plot: Third Period Grade by Absences") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title
```

What's Interesting?: More females then males were absent.

7) Repeat the scatter plot step but use faceting to obtain graphs at each setting of another categorical variable.

```{r}
# Scatter Plot with faceting: G3 and absences across internet

# Setting up labels for facet plots
custom_labels <- c("yes" = "Internet Access: Yes", "no" = "Internet Access: No")

sp1 <- ggplot(student_data |> drop_na(G3, absences), aes(x = absences, y = G3, color = internet))
sp1 + geom_point() +
  labs(x = "Absences", y = "Third Period Grades", 
       title = "Scatter Plot: Third Period Grade by Absences, Sex, and Internet Access") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ internet, labeller = as_labeller(custom_labels)) 
```

What's Interesting?: In the third period, there are more males then females with internet access.

8) Repeat the scatter plot step but use faceting to obtain graphs at each combination of two categorical variables.

```{r}
# Scatter Plot with faceting: G3 and absences across sex internet

# Setting up labels for facet plots
custom_labels <- c("yes" = "Internet Access: Yes", "no" = "Internet Access: No")

sp1 <- ggplot(student_data |> drop_na(G3, absences), aes(x = absences, y = G3, color = sex))
sp1 + geom_point() +
  labs(x = "Absences", y = "Third Period Grades", 
       title = "Scatter Plot: Third Period Grade by Absences, Sex, and Internet Access") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ internet, labeller = as_labeller(custom_labels)) 
```

What's Interesting?: In the third period, the proportion of student absences is the same regarding internet access.